---
title: "Visualising read-level methylation with ggmethylation"
author: "Luuk Harbers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising read-level methylation with ggmethylation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
cache_path <- system.file("extdata", "vignette_cache.rds",
                           package = "ggmethylation")
has_cache  <- nzchar(cache_path) && file.exists(cache_path)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  eval = has_cache
)
```

```{r load-cache, include = FALSE}
library(ggmethylation)
library(ggplot2)

cache       <- readRDS(cache_path)
meth_basic  <- cache$meth_basic
meth_6ma    <- cache$meth_6ma
meth_hp     <- cache$meth_hp
meth_snv    <- cache$meth_snv
variants    <- cache$variants
annotations <- cache$annotations
region      <- cache$region
```

## Introduction

`ggmethylation` visualises read-level base modification data from long-read
sequencing (Oxford Nanopore, PacBio). It parses the MM/ML auxiliary tags
written by basecallers and tools such as `dorado`, `modkit` or `pb-CpG-tools`,
and produces a composite `patchwork` figure:

- **Top panel** — individual reads drawn as horizontal bars, with each
  modification site coloured by its per-site probability.
- **Bottom panel** — a loess-smoothed summary of modification probability,
  one curve per group.

Optional panels sit above and below these: a gene annotation track, and a
group-difference (delta) track.

Reads can be grouped by any BAM auxiliary tag (typically the haplotype tag
`HP`) or by the genotype at a supplied SNV position. Grouping drives every
panel, so the relationship between haplotype and methylation state is directly
readable off the figure.

**Index requirement.** `ggmethylation` uses `Rsamtools` to query reads by
region, so a BAM index (`.bai`, from `samtools index`) must sit alongside the
BAM before you call `read_methylation()`.

### The data used in this vignette

Every figure below is built from the package's test fixture:

```{r fixture-info, echo = FALSE}
cat("Region: ", region, "\n", sep = "")
cat("Reads:  ", nrow(meth_basic$reads), "\n", sep = "")
cat("Sites:  ", nrow(meth_basic$sites), " (5mC)\n", sep = "")
```

This is HG002 PacBio Fiber-seq data, haplotype-phased with `longphase` and
aligned to CHM13, subset to the **MEG3** promoter. MEG3 is an imprinted
lncRNA: one parental allele is methylated at the promoter and the other is
not. That makes it a good worked example, because the grouped figures show a
genuine biological difference rather than noise.

## Installation

```{r installation, eval = FALSE}
# Bioconductor dependencies
BiocManager::install(c(
  "Rsamtools",
  "GenomicRanges",
  "GenomicAlignments",
  "IRanges"
))

# Install ggmethylation from GitHub
devtools::install_github("ljwharbers/ggmethylation")
```

## Reading data

`read_methylation()` fetches all reads overlapping a region and returns a
`methylation_data` S3 object of tidy data frames.

```{r read-basic-show, eval = FALSE}
library(ggmethylation)

meth_basic <- read_methylation(
  "path/to/sample.bam",
  "chr14:95055000-95070000",
  mod_code = "m"          # 5mC; use "a" for 6mA
)
```

```{r print-basic}
meth_basic
```

The object holds:

| Element | Contents |
|---|---|
| `$reads` | one row **per read**: `read_name`, `start`, `end`, `strand`, `mean_mod_prob`, and `group` when grouped |
| `$sites` | one row **per modification call per read**: `read_name`, `position`, `mod_prob`, `mod_code` |
| `$insertion_sites` | calls on inserted bases, which have no reference coordinate |
| `$cigar_features` | one row per CIGAR operation per read |
| `$region`, `$mod_code`, `$group_tag` | the query metadata |

```{r structure}
head(meth_basic$reads, 3)
head(meth_basic$sites, 3)
```

Note that `$sites` does **not** contain every canonical base for every read.
`?read_methylation` documents the five reasons a position can be absent — the
MM tag's `?`/`.` flag, basecaller convention, sequence differences, CIGAR
effects, and differing alignment spans.

### Useful filters

```{r read-filters, eval = FALSE}
read_methylation(
  bam, region,
  mod_code      = "m",
  min_mapq      = 20,      # drop poorly-mapped reads
  strand        = "+",     # keep one strand only
  min_length    = 5000,    # drop short reads
  max_reads     = 200      # downsample for very deep regions
)
```

## The basic plot

```{r plot-basic}
plot_methylation(meth_basic)
```

Reads are packed into lanes like a genome browser. Each coloured tick is one
5mC call, running from `colour_low` (unmethylated) to `colour_high`
(methylated).

## Grouping by haplotype

Passing `group_tag` splits the reads on a BAM tag. `drop_na_group = FALSE`
keeps reads that carry no tag as a separate `NA` group.

```{r read-hp-show, eval = FALSE}
meth_hp <- read_methylation(
  bam, region,
  mod_code      = "m",
  group_tag     = "HP",
  drop_na_group = TRUE
)
```

```{r plot-hp, fig.height = 6}
plot_methylation(meth_hp)
```

The imprinting is now obvious: one haplotype is methylated across the MEG3
promoter and the other is not. Quantitatively:

```{r hp-quant}
tss  <- 95060996
near <- meth_hp$sites[abs(meth_hp$sites$position - tss) < 2000, ]
round(tapply(near$mod_prob, near$group, mean), 3)
```

## Confidence ribbon on the smooth

`show_ci` (on by default) draws the loess confidence interval behind each
curve, so you can see where the smooth is well supported and where it is
resting on a handful of reads.

```{r show-ci, fig.height = 6}
plot_methylation(meth_hp, show_ci = TRUE)
```

Set `show_ci = FALSE` to get the bare lines back.

## The group-difference (delta) track

When the data splits into **exactly two** groups, `show_delta = TRUE` appends
a panel showing the signed difference between them, so you read *where* the
groups diverge instead of eyeballing the gap between two curves.

```{r show-delta, fig.height = 7}
plot_methylation(meth_hp, show_delta = TRUE)
```

The area is filled with the colour of whichever group is higher at each
position, so the sign is readable directly off the plot. With any number of
groups other than two, the panel is skipped with a warning.

## Binary call mode

By default each site is coloured on a continuous probability gradient.
`call_mode = "binary"` instead renders discrete calls, which cuts visual noise
when you care about the call rather than the score.

```{r call-mode, fig.height = 6}
plot_methylation(meth_hp, call_mode = "binary", call_threshold = 0.5)
```

`call_ambiguous` adds a third state for low-confidence calls: any site within
that half-width of the threshold is drawn in `colour_ambiguous` and excluded
from the binary aggregate.

```{r call-ambiguous, fig.height = 6}
plot_methylation(
  meth_hp,
  call_mode      = "binary",
  call_threshold = 0.5,
  call_ambiguous = 0.2      # 0.3-0.7 becomes "ambiguous"
)
```

In binary mode the smooth panel switches from a mean probability to the
**fraction of calls that are methylated**, and its axis label changes to match.

## Grouping by SNV genotype

Instead of a tag, reads can be split by the base they carry at a given
position. This works on unphased data, and is a good check that an
allele-specific signal is real.

```{r read-snv-show, eval = FALSE}
meth_snv <- read_methylation(
  bam, region,
  mod_code     = "m",
  snv_position = 95066012,
  ref_base     = "C",
  alt_base     = "T"
)
```

```{r plot-snv, fig.height = 6}
table(meth_snv$reads$group)
plot_methylation(meth_snv)
```

Reads carrying neither the reference nor the alternate base are dropped.

## Gene annotation track

`read_annotations()` fetches a gene model for the region. Supply `genome`
(`"hg38"` or `"chm13"`) to download and cache the UCSC ncbiRefSeq GTF, or pass
your own `gtf`/`txdb`.

```{r annot-show, eval = FALSE}
annotations <- read_annotations(genome = "chm13", region = region)
```

```{r plot-annot, fig.height = 7}
plot_methylation(meth_hp, annotations = annotations)
```

The gene panel is placed at the top and shares the genomic x-axis with every
other panel. `clear_annotation_cache()` drops the cached TxDb if you need to
force a rebuild.

## Variant overlays

`read_variants()` reads a VCF and classifies each record as an SNV, an
insertion, a deletion, or a structural variant (`DEL`, `DUP`, `INV`, `BND`).
Passing the result to `plot_methylation()` overlays them on the read panel.

```{r variants-show, eval = FALSE}
variants <- read_variants("path/to/calls.vcf.gz", region)
```

```{r print-variants}
variants
```

```{r plot-variants, fig.height = 6}
plot_methylation(meth_hp, variants = variants)
```

SNVs are marked with an asterisk **only on the reads that actually carry the
alternate allele**, which is why the marks track the haplotypes here. Because
that requires the read sequences, the object must come from a current
`read_methylation()` call.

Structural variants are drawn as spanning bars, and BND records are labelled
with their mate location. `bnd_match_tol` controls how far a read's
supplementary alignment may sit from a VCF BND position and still be
considered a match.

## CIGAR features and supplementary alignments

`show_cigar = TRUE` (the default) marks large indels taken from the CIGAR
string; `min_indel_size` sets the threshold. `show_supplementary = TRUE` adds
arrowheads to reads with supplementary alignments, flagging candidate
structural rearrangements.

```{r cigar, fig.height = 6}
plot_methylation(
  meth_hp,
  show_cigar         = TRUE,
  min_indel_size     = 50,
  show_supplementary = TRUE
)
```

Consensus deletions also break the smoothed curve and its confidence ribbon,
so the summary line is not interpolated across a region where the reads carry
no sequence.

## Multiple modifications at once

Fiber-seq and similar protocols call more than one modification. `mod_code`
selects which to read:

```{r multi-mod}
c(`5mC sites` = nrow(meth_basic$sites),
  `6mA sites` = nrow(meth_6ma$sites))
```

```{r plot-6ma, fig.height = 5}
plot_methylation(meth_6ma)
```

When an object carries several modification codes, `mod_code_shapes` maps each
code to its own point shape so they stay distinguishable in one panel.

## Insertion-aware modification calls

Modifications on inserted bases have no reference coordinate, so they cannot
appear in the main plot. They are kept in `$insertion_sites`, and
`list_insertion_loci()` clusters insertion events across reads into loci.

```{r insertion-loci}
insertion_sites(meth_basic) |> head(3)

loci <- list_insertion_loci(meth_basic, tol_pos = 10L, tol_len = 0.20,
                            min_reads = 2L)
head(loci[, c("locus_id", "anchor_pos", "n_carriers", "n_noncarriers")], 5)
```

`plot_insertion_locus()` then visualises one locus in a stitched coordinate
system — left reference flank | inserted bases | right reference flank —
comparing reads that carry the insertion against those that do not. Use the
`locus_id` column above to pick one.

```{r plot-insertion, fig.height = 5}
plot_insertion_locus(meth_basic, loci$locus_id[1], flank = 200)
```

This makes it possible to ask whether an insertion changes the modification
state of the sequence around it — chromatin effects on nearby 6mA, for
instance. `include_noncarriers = FALSE` restricts the panel to carriers, and
`show_smoothed = FALSE` returns the read panel alone as a plain ggplot.

## Multiple samples

`merge_methylation()` combines several `methylation_data` objects covering the
same region into one `multi_methylation_data` object, which
`plot_methylation()` renders as stacked per-sample read panels above a shared
smooth panel.

```{r multi-sample-show, eval = FALSE}
md1    <- read_methylation("tumour.bam", region, mod_code = "m")
md2    <- read_methylation("normal.bam", region, mod_code = "m")
merged <- merge_methylation(tumour = md1, normal = md2)
plot_methylation(merged)
```

The fixture here is a single sample, so to demonstrate the layout we split its
two haplotypes into two pseudo-samples. This is a presentational trick for the
vignette, not an analysis you would run:

```{r multi-sample, fig.height = 7}
split_by_group <- function(m, g) {
  keep      <- m$reads$read_name[m$reads$group == g]
  m$reads   <- m$reads[m$reads$group == g, ]
  m$sites   <- m$sites[m$sites$read_name %in% keep, ]
  m$reads$group <- NULL
  m$sites$group <- NULL
  m$group_tag   <- NULL
  m
}

merged <- merge_methylation(
  hap1 = split_by_group(meth_hp, "1"),
  hap2 = split_by_group(meth_hp, "2")
)
merged
plot_methylation(merged)
```

## Customising appearance

```{r custom-colours, fig.height = 6}
plot_methylation(
  meth_hp,
  colour_low  = "#F7F7F7",
  colour_high = "#2166AC",
  line_width  = 0.4          # thickness of the read bars
)
```

`group_colours` sets the per-group palette (the default is a colourblind-safe
Okabe–Ito pair), and `strand_colours` applies when `colour_strand = TRUE`.

Relative panel heights are set with `panel_heights`, whose expected length
depends on how many optional panels are switched on:

```{r custom-panels, fig.height = 6}
plot_methylation(meth_hp, panel_heights = c(2, 1))
```

Reads can be sorted by any column of `$reads`:

```{r custom-sort, fig.height = 6}
plot_methylation(meth_hp, sort_by = "mean_mod_prob")
```

Valid values include `"start"`, `"end"`, `"strand"`, `"group"` and
`"mean_mod_prob"`. An unknown column raises an error naming the columns that
are available — note that `"position"` belongs to `$sites`, not `$reads`.

`theme_ggmethylation()` exposes the shared panel theme if you want to build
panels yourself or restyle the output.

## Exporting

`write_methylation()` writes the reads and sites tables to TSV or BED,
optionally gzipped. It returns its input invisibly, so it slots into a
pipeline.

```{r write, eval = FALSE}
write_methylation(meth_hp, prefix = "output/meg3")                  # TSV
write_methylation(meth_hp, prefix = "output/meg3",
                  format = "bed", gzip = TRUE)                      # BED + gzip
write_methylation(meth_hp, prefix = "output/meg3", sites = FALSE)   # reads only
```

BED output is 0-based half-open, with `mod_prob` in the score column.

To save a figure, treat it as any other ggplot/patchwork object:

```{r ggsave, eval = FALSE}
p <- plot_methylation(meth_hp, show_delta = TRUE)
ggsave("meg3.png", p, width = 9, height = 7, dpi = 300)
```

## Session info

```{r session-info, eval = TRUE, echo = TRUE}
sessionInfo()
```
