---
title: "Visualising read-level methylation with ggmethylation"
author: "Luuk Harbers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising read-level methylation with ggmethylation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Introduction

`ggmethylation` provides read-level visualisation of 5-methylcytosine (5mC)
from long-read sequencing data stored in modBAM format. It parses the MM/ML
auxiliary tags written by tools such as pb-CpG-tools or modkit and produces
a two-panel `patchwork` figure:

- **Top panel** — per-group methylation summary (mean smoothed curve or bar).
- **Bottom panel** — individual read methylation tiles, coloured by per-site
  modification probability.

Reads can be grouped by any BAM auxiliary tag (e.g. haplotype tag `HP`) or by
the genotype at a user-supplied SNV position. Grouping drives both panels
independently, so the relationship between haplotype/genotype and methylation
state is immediately apparent.

**Index requirement.** `ggmethylation` uses `Rsamtools` to query reads by
genomic region. A BAM index (`.bai` file, produced with `samtools index`) must
be present alongside the BAM file before calling `read_methylation()`.

## Installation

```{r installation, eval = FALSE}
# Bioconductor dependencies
BiocManager::install(c(
  "Rsamtools",
  "GenomicRanges",
  "GenomicAlignments",
  "IRanges"
))

# Install ggmethylation from GitHub
devtools::install_github("lharbers/ggmethylation")
```

## Basic usage: ungrouped methylation plot

`read_methylation()` fetches all reads overlapping a genomic region and
returns a `methylation_data` S3 object containing tidy data frames ready for
plotting.

```{r read-basic, eval = FALSE, echo = TRUE}
library(ggmethylation)

bam    <- "path/to/sample.bam"
region <- "chr21:34500000-34510000"

meth_basic <- read_methylation(bam, region)
```

```{r load-cache, eval = TRUE, echo = FALSE}
cache      <- readRDS(system.file("extdata", "vignette_cache.rds",
                                  package = "ggmethylation"))
meth_basic <- cache$meth_basic
```

Printing the object gives a concise summary of the region, read count, and
available grouping:

```{r print-basic, eval = TRUE, echo = TRUE}
print(meth_basic)
```

Pass the object directly to `plot_methylation()` to produce the two-panel
figure:

```{r plot-basic, eval = TRUE, echo = TRUE}
library(ggmethylation)
plot_methylation(meth_basic)
```

## Accessing the methylation_data object

The `methylation_data` object is a named list. The two most useful elements
are `$reads` (one row per read per CpG site) and `$sites` (genomic-coordinate
summary).

```{r str-basic, eval = TRUE, echo = TRUE}
str(meth_basic, max.level = 1)
```

```{r head-reads, eval = TRUE, echo = TRUE}
head(meth_basic$reads)
```

```{r head-sites, eval = TRUE, echo = TRUE}
head(meth_basic$sites)
```

## Haplotype-based grouping (HP tag)

PacBio HiFi data processed with DeepVariant + WhatsHap carries a per-read
haplotype tag (`HP`). Supply `group_tag = "HP"` to stratify the plot by
haplotype.

```{r read-hp, eval = FALSE, echo = TRUE}
meth_hp <- read_methylation(bam, region, group_tag = "HP")
```

```{r load-hp, eval = TRUE, echo = FALSE}
meth_hp <- cache$meth_hp
```

```{r plot-hp, eval = TRUE, echo = TRUE}
plot_methylation(meth_hp)
```

Group colours can be overridden with a named character vector passed to
`group_colours`. Names must match the levels returned by the HP tag.

```{r plot-hp-colours, eval = TRUE, echo = TRUE}
plot_methylation(
  meth_hp,
  group_colours = c("1" = "steelblue", "2" = "coral")
)
```

## SNV-based grouping

When a heterozygous SNV is present in the region, reads can be split by
allele using `snv_position`, `ref_base`, and `alt_base`. A dashed vertical
line marks the SNV position in the read panel so the allele split is easy to
locate visually.

```{r read-snv, eval = FALSE, echo = TRUE}
# Replace snv_position with the actual coordinate of a heterozygous SNV
meth_snv <- read_methylation(
  bam, region,
  snv_position = 34505000L,
  ref_base     = "C",
  alt_base     = "T"
)
```

```{r load-snv, eval = TRUE, echo = FALSE}
meth_snv <- cache$meth_snv
```

```{r plot-snv, eval = TRUE, echo = TRUE}
plot_methylation(meth_snv)
```

The dashed vertical line in the lower panel marks the SNV position and
visually anchors the two allele groups. In the top panel, the mean methylation
curves for REF and ALT reads are overlaid for direct comparison.

## Customising plot appearance

`plot_methylation()` exposes several arguments for fine-tuning the figure.
The examples below use `meth_hp` from the cache.

```{r custom-colours, eval = TRUE, echo = TRUE}
# Custom methylation colour gradient
plot_methylation(
  meth_hp,
  colour_low  = "white",
  colour_high = "darkgreen"
)
```

```{r custom-dots, eval = TRUE, echo = TRUE}
# Larger dots and a wider smoother
plot_methylation(
  meth_hp,
  dot_size    = 2.5,
  smooth_span = 0.4
)
```

```{r custom-panels, eval = TRUE, echo = TRUE}
# Taller summary panel relative to read panel
plot_methylation(
  meth_hp,
  panel_heights = c(2, 3)
)
```

```{r custom-sort, eval = TRUE, echo = TRUE}
# Sort reads by group membership (default) vs. by genomic start
plot_methylation(meth_hp, sort_by = "position")
```

## Composing with ggplot2

`plot_methylation()` returns a `patchwork` object, which can be modified with
standard ggplot2 theme calls applied via `&` (applies to all panels) or `+`
(applies to the last panel only).

```{r ggplot2-compose, eval = TRUE, echo = TRUE}
library(ggplot2)

p <- plot_methylation(meth_basic)

# Increase base font size across both panels
p & theme(text = element_text(size = 14))
```

Save to file with `ggplot2::ggsave()`. Supply explicit `width` and `height`
because patchwork objects do not carry intrinsic dimensions.

```{r ggsave, eval = FALSE, echo = TRUE}
ggsave("my_methylation_plot.pdf", p, width = 10, height = 6)
```

Patchwork operators (`/`, `|`, `+`) can be used to combine a
`plot_methylation()` output with other ggplot2 panels in a larger figure
layout.

## Session info

```{r session-info, eval = TRUE, echo = TRUE}
sessionInfo()
```
